---
title: "Cleaning and Processing"
output: html_notebook
---


```{r setup, include=FALSE}
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#install and load packages
```{r install packages}
library(skimr)
library(janitor)
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyr)

```
#importing the datasets
Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.
```{r import datasets}
activity <- read.csv("../Fitbit/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
calories <- read.csv("../Fitbit/Fitabase Data 4.12.16-5.12.16/hourlyCalories_merged.csv")
intensities <- read.csv("../Fitbit/Fitabase Data 4.12.16-5.12.16/hourlyIntensities_merged.csv")
sleep <- read.csv("../Fitbit/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
weight <- read.csv("../Fitbit/Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")
heart <- read.csv("../Fitbit/Fitabase Data 4.12.16-5.12.16/heartrate_seconds_merged.csv")
```
#a look at the datasets

```{r}
glimpse(activity)
head(activity)
glimpse(intensities)
head(intensities)

```
I spotted some problems with the timestamp data. So before analysis, I need to convert it to date time format and split to date and time.
#Harmonizng date formating
```{r}
#activity
activity$ActivityDate=as.POSIXct(activity$ActivityDate, format="%m/%d/%Y", tz=Sys.timezone())
activity$date = format(activity$ActivityDate, format="%m/%d/%Y")

#calories
calories$ActivityHour=as.POSIXct(calories$ActivityHour, format="%m/%d/%Y %I:%M:%S %p ", tz=Sys.timezone())
calories$date = format(calories$ActivityHour, format="%m/%d/%Y")
calories$time = format(calories$ActivityHour , format ="%H:%M:%S")

#heart
heart$Time = as.POSIXct(heart$Time, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
heart$date = format(heart$Time, format="%m/%d/%Y" )
heart$time = format(heart$Time, format = "%H:%M:%S")

# intensities
intensities$ActivityHour=as.POSIXct(intensities$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
intensities$time <- format(intensities$ActivityHour, format = "%H:%M:%S")
intensities$date <- format(intensities$ActivityHour, format = "%m/%d/%Y")

# sleep
sleep$SleepDay=as.POSIXct(sleep$SleepDay, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
sleep$date <- format(sleep$SleepDay, format = "%m/%d/%Y")
sleep$time = format(sleep$SleepDay, format ="%H:%M:%S" )
```
exploring the datasets
```{r rnning some exploration}
#check the number of distinct participants ID
n_distinct(activity$Id)
n_distinct(calories$Id)
n_distinct(intensities$Id)
n_distinct(sleep$Id)
n_distinct(weight$Id)

```
this tells us that there were 
-33 participants who provided their data for the activities,calories,intensities data sets.
-24 participants who provided their data for the sleep data set.
-8 participants who provided their data for the weight data set. This sample is too low, the weight data set cannot be used to make any meaningful insights.

Now we run some summary statistics
```{r summary statistics}
# activity
activity %>%  
  select(TotalSteps,
         TotalDistance,
         SedentaryMinutes, Calories) %>%
  summary()
# explore num of active minutes per category
activity %>%
  select(VeryActiveMinutes, FairlyActiveMinutes, LightlyActiveMinutes) %>%
  summary()
# calories
calories %>%
  select(Calories) %>%
  summary()
#heart
heart %>%
  select(Value) %>%
  summary
# sleep
sleep %>%
  select(TotalSleepRecords, TotalMinutesAsleep, TotalTimeInBed) %>%
  summary()
# weight
weight %>%
  select(WeightKg, BMI) %>%
  summary()
#intensities
intensities %>%
  select(TotalIntensity,AverageIntensity) %>%
  summary()

```

#some interesting findings
-the average sedentary minutes was 991.2 that is more than 16 hours. That is not good
-the average participant weight was 72kg with a BMI of 24.19, so the participants are quite fit
-on average participants sleep once every 7 hours
-Average total steps per day are 7638 which a little bit less for having health benefits for according to the CDC research. They found that taking 8,000 steps per day was associated with a 51% lower risk for all-cause mortality (or death from all causes). Taking 12,000 steps per day was associated with a 65% lower risk compared with taking 4,000 steps.
#Merging data
Before beginning to visualize the data, We will going to merge (inner join) activity and sleep on columns Id and date 
```{r}

```
## Plotting a few explorations

What's the relationship between steps taken in a day and sedentary minutes? How could this help inform the customer segments that we can market to? E.g. position this more as a way to get started in walking more? Or to measure steps that you're already taking?
```{r total steps and calories}
ggplot(activity, aes(x=TotalSteps, y=Calories)) + geom_point() + geom_smooth() + labs(title="Total Steps vs. Calories")
```
We see positive correlation here between Total Steps and Calories, which is obvious - the more active we are, the more calories we burn.

```{r VeryActiveMinutes and Calories}
ggplot(activity, aes(x=VeryActiveMinutes, y=Calories))+geom_point() + geom_smooth() + labs(title="Activity vs. Calories")
```
I see positive correlation here between VeryActiveMinutes and Calories, which is obvious - the longer and more active we are, the more calories we burn.
```{r}
ggplot(data=sleep, aes(x=TotalMinutesAsleep, y=TotalTimeInBed)) + 
  geom_point()+ labs(title="Total Minutes Asleep vs. Total Time in Bed")
```
The relationship between Total Minutes Asleep and Total Time in Bed looks linear. *We could set up notifications to remind Bellabeat users to go to bed early in order to improve their sleep quality*

Let's look at intensities data over time (hourly).
```{r create a new dataframe with time and the average intensity for each hour of the day}
int_new <- intensities %>%
  group_by(time) %>%
  drop_na() %>%
  summarise(mean_total_int = mean(TotalIntensity))

ggplot(data=int_new, aes(x=time, y=mean_total_int)) + geom_histogram(stat = "identity", fill='darkblue') +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title="Average Total Intensity vs. Time")

```
-participants were most active between 5 and 12pm, and between 5 and 7pm.*This suggest that the participants are working full time jobs*
-the highest range of intensity ie: 5-7pm could be de to participants going to a gym or to a walk after closing from work.We can use this time in the Bellabeat app to remind and motivate users to go for a run or walk.
```{r merging}
merged_data <- merge(sleep, activity, by=c('Id', 'date'))
head(merged_data)
```
```{r mergin activity,sleep,heart}

```

```{r}
ggplot(data=merged_data, aes(x=TotalMinutesAsleep, y=SedentaryMinutes)) + 
geom_point(color='darkblue') + geom_smooth() +
  labs(title="Minutes Asleep vs. Sedentary Minutes")
```

-negative relationship between sedentary minutes and total minutes asleep

-As an idea: *if Bellabeat users want to improve their sleep, Bellabeat app can recommend reducing sedentary time.*
```{r}
#add a day_of_week column to intensities data set
intensities$Day_of_week = wday(intensities$ActivityHour,label=TRUE, abbr=FALSE)
#create a new data frame 
int_new_wday <- intensities %>%
  group_by(Day_of_week) %>%
  drop_na() %>%
  summarise(mean_total_int_wday = mean(TotalIntensity))
#plot week_day and and average intensity
ggplot(data=int_new_wday, aes(x=Day_of_week, y=mean_total_int_wday)) + geom_histogram(stat = "identity", fill='darkblue') +
  labs(title="Average Total Intensity vs. Weekday")

```
```{r export the datasets for vizalisation in tableau}
write.csv(activity, "../Fitbit/Fitabase Data 4.12.16-5.12.16/new_activity.csv", row.names = FALSE)

write.csv(merged_data, "../Fitbit/Fitabase Data 4.12.16-5.12.16/merged_data.csv", row.names = FALSE)

write.csv(int_new, "../Fitbit/Fitabase Data 4.12.16-5.12.16/int_new.csv", row.names = FALSE)
 
write.csv(int_new_wday, "../Fitbit/Fitabase Data 4.12.16-5.12.16/int_new_wday.csv", row.names = FALSE)
```

